---
title: "Tutorial 6.2: Linear regression (continuous) on the fish tank dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

As an exercise on linear regression, we will analyse
the fish tank dataset.

# Fish poison dataset

In this experiments 96 fish (dojofish, goldfish and zebrafish)
were placed separately in a tank with two liters of water and
a certain dose (in mg) of a certain poison EI-43,064. The resistance
of the fish a against the poison was measured as the amount of
minutes the fish survived upon adding the poison (Surv_time, in
minutes). Additionally, the weightt of each fish was measured.

# Goal

In this tutorial, we will study the association between dose and
survival time by using a linear regression model.

Read the required libraries

```{r, message = FALSE}

```

# Import the data

```{r, message=FALSE}
poison <- read_csv("https://raw.githubusercontent.com/GTPB/PSLS20/master/data/poison.csv")
```

# Data tidying

Have a look at the data.

We can see a couple of things in the data that can
be improved upon:

1. Capitalize the fist column name 
2. Set the Species column as a factor
3. Change the speciec factor levels from 0, 1 and 2 to
Dojofish, Goldfish and Zebrafish. **Hint**: use the `fct_recode`
function.

```{r}

```


# Data Exploration

Make a suitable visualization of the association between
the dose and the survival time. **Hint:** put Dose on the
x-axis and survival time on the y-axis. Try to see if
a straight line can capture the trend in the data.

```{r}

```

How many fish do we have per species?

```{r}

```


# Assumptions of linear regression

List assumptions:

1. The observations are independent of each other
2. Linearity between the response and predictor variable
3. The residues of the model must be normally distributed
4. Homoscedasticity of the data

## Independence

The first assumption is **not met**. There are patterns
of relatedness in the data. Indeed, we expect that
fish of the same species have a more similar response to
the poison as compared to fish of another species.
Equivalently, we expect that fish of the same weight have 
a more similar response to the poison as compared to fish 
of different weight. 

** For now, we will ignore this problem **. In a later 
tutorial, we will see how we can account for this dependence
by including the species data and weigth data in the linear model
(class on multiple regression). As such, ** you may assume **
** the assumption is met. **

We will check the other assumptions by first fitting the linear
model and plotting the output. As such, we will get all the
required diagnostic plots.

```{r}
model <- lm(...~..., data=...) 

## display the diagnostic plots of the model
plot(model)
```

We have four diagnostic plots:

## Linearity with the Residuals vs fitted plot

- predictor of predictions $\hat\beta_0+\hat\beta_1 x$ on $X$-axis
- *residuals* on $Y$-as
$$e_i=y_i-\hat{g}(x_i)=y_i-\hat\beta_0-\hat\beta_1\times x_i,$$

If there would be a linear relationship in the data,
the residuals are expected to lie on the y=0 line for
the entire range of predicted values. 

Based on this explanation, is the 2nd assumption met?

## Normal Q-Q

- QQ-plot of the residuals $e_i$.

The residuals of the linear regression model should be normally
distributed. Do you think the assumption is met?

## Homoscedasticity

- Square-root of the absolute value of standardized residuals
in function of the fitted values

To meet the third assumption of linear regression, the variance
on the _Square-root of the absolute value of standardized residuals_
must be similar over the entire range of fitted values. The smoother
in the plot helps us with looking at this; it should be nicely
horizontal over the entire range of fitted values. 

Do you think this assumption is met?

# Log transormation

Multiple assumptions are not met, so way not continue by performing
a linear regression.

One possibility is to log-transform the data. Indeed, data with heavy
right tails can often be normalized by applying this strategy

Generate a new column in the poison dataset, containing the 
log-transformed survival times.

```{r}
poison <- poison %>%
  mutate(log.Surv_time = ...)
```

Refit the (log-)linear model. Generate diagnostic plots.
Assess the assumptions.

```{r}
log.model <- ...
plot(log.model)
```

Upon log-transformation, are all the required assumptions met?

Look at the output of the log-linear model:

```{r}
summary(log.model)
```

Compute the 95% confidence interval on the model parameters:

```{r}
confint(log.model)
```

## Interpretation on the log scale 

Currently, all the outcomes should be interpreted on the log-scale.
Indeed, since we are now modelling  _the log of survival time_, we 
don't have direct inference on survival time.

As such, we may interpret the output as follows:

- The slope:

If the dose at which the poison is administered increases with 1 mg,
then the mean of the natural logarithm of survival time will decrease
by 0.5112 units (95% CI [-0.8109,-0.2116]). This decrease is significant
on the 5% significance level (p=0.00103).

- The intercept:

The average of the natural logarithm of survival time for fish that
were given a dose of 0 mg is 2.1046 (95% CI [1.6288, 2.5804]). This
value is significantly different from 0 on the 5% significance level
(7.13e-14). This conclusion will be discussed in more detail below.

## Interpretation on the original scale 

The interpretation the log-scale is quite difficult: the natural 
logarithm survival time is not exactly a casual measure. To ease 
the interpetation, we may backtransform the results to the original
scale (time in minutes). This we can do by taking the exponent of
the outcomes:

```{r}
exp(summary(log.model)$coefficients[,"Estimate"])
```

and of their confidence intervals:

```{r}
exp(confint(log.model))
```

Now, we can interpret the results in terms of the geometric mean:

- The slope:

If the dose at which the poison is administered increases with 1 mg,
then the geometric mean of survival time will decrease by a factor
of 0.5998 (95% CI [0.4445,0.8093]). This decrease is significant
on the 5% significance level (p=0.00103).

- The intercept:

The geometric mean of the survival time for fish that were given a 
dose of 0 mg is 8.2039 minutes (95% CI [5.0977, 13.2029]). This
value is significantly different from 0 on the 5% significance level
(7.13e-14).

To go a little further on the latter interpretation, our analysis seems
to suggest that fish in uncontaminated water would only suvive for about
8 minutes! This seems a little bit _fishy_...

Here, we make a diagnostic plot to explain this behavior:

```{r}
plot(x=poison$Dose, y=poison$Surv_time, xlab="Dose (mg)", ylab="Survival time (min)", xlim=c(0,2.2))

abline(lsfit(poison$Dose,poison$Surv_time), lty=2) ## lty=2 draws a dashed line

## add the intercept as a point
summary_model <- summary(model)
points(0,summary_model$coefficients["(Intercept)","Estimate"],col="red", pch=16)
```

We can see that the dose value of 0 mg does not fall within
the same range as our data. As such, it is unlikely that we
can assume this model to have correct inference on the intercept:
we expect that the linearity assumption will not be met for these
low values. The model will only be trustworthy on the range of
data that we use to build the model on!

This is not really a problem: we are mainly interested in the slope
parameter in the given range of datapoints.

## Conclusion

- On the log scale:

If the dose at which the poison is administered increases with 1 mg,
then the mean of the natural logarithm of survival time will decrease
by 0.5112 units (95% CI [-0.8109,-0.2116]). This decrease is highly
significant on the 5% significance level (p=0.00103).

- On the original scale:

If the dose at which the poison is administered increases with 1 mg,
then the geometric mean of survival time will decrease by a factor
of 0.5998 (95% CI [0.4445,0.8093]). This decrease is highly significant
on the 5% significance level (p=0.00103).









