---
title: "9. Nonparametric Statistics - Wilcoxon-Mann-Withney test"  
author: "Lieven Clement"
date: "statOmics, Ghent University (https://statomics.github.io)"
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, comment = NA, echo = TRUE,
                      message = FALSE, warning = FALSE)
library(tidyverse)
library(Rmisc)
set.seed(140)
```


#Intro

Inference was only correct if distributional assumptions were satisfied 

- e.g Normal distribution 
- equal variance 
 
-  The $p$-value: $\text{P}_0\left[ \vert T\vert \geq \vert t \vert \right]$. 
  	
	- Calculated using the null distribution of $T$ that we derived under the assumptions
	- In correct if assumptions are violated  
	
-  $95\%$ CI also builds upon these assumptions. If they are invalid then the intervals will not contain the population parameter with 95% probability.

- Asymptotic theory is more difficult to place: the $t$-test ais symptotically non-parametric because for very large samples the distributional assumptions of normality are no longer important. 

- If assumptions hold the parametric approach

	- more efficient: larger power with same sample size + smaller CI.  
	- more flexible: easier to analyse data with complex designs  
	

## Cholestorol voorbeeld

- Cholestorol concentration in blood measured for 
  - 5 patients (groep=1) two days upon a stroke
  - 5 healty subject (groep=2). 

- Is cholestorol concentration of hartpatients and healthy subjects on average different? 

```{r}
chol <- read_tsv("https://raw.githubusercontent.com/GTPB/PSLS20/master/data/chol.txt")
chol$group <- as.factor(chol$group)
nGroups <- table(chol$group)
n <- sum(nGroups)
chol
```

---


```{r, echo=FALSE, fig.align='center'}
chol %>%  ggplot(aes(x=group,y=cholest)) + geom_boxplot(outlier.shape=NA) + geom_point(position="jitter")

chol %>% ggplot(aes(sample=cholest)) + 
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~group) 
```

- Possibly outliers
- Difficult to assess distributional assumptions when only 5 observations are available. 


#Rank Tests

- Important group of non-parametric test
  - Non-parametric, 
  - Exact $p$-values using a permutatie null distribution. 
  - No need for separate permutation distribution for each new dataset.  
  - Permutation null distribution of rank tests only depends on sample size 
  - Robust to outliers 
  
#Ranks

Rank tests start from rank-transformed data. 

- Let $Y_1, \ldots, Y_n$. 
- In the absence of *ties* 
  $$R_i=R(Y_i) = \#\{Y_j: Y_j\leq Y_i; j=1,\ldots, n\}$$
- Smallest observation has rank 1, second smallest rank 2, ... , largest observation gets rank $n$ 

```{r}
chol$cholest
rank(chol$cholest)
```

## ties 

Sometimes *ties* occur: two observations with identical values 

```{r}
withTies=c(403,507,507,610,651,651,651,830,900)
rank(withTies)
```

- Ties: 507 occurs twice, 651 occurs 3 times
- If ties occur *midranks* are used. 

- **midrank** of observation $Y_i$ becomes
  \begin{eqnarray*}
   R_i &=& \frac{ \#\{Y_j: Y_j\leq Y_i\} + ( \#\{Y_j: Y_j < Y_i\} +1)}{2}.
   \end{eqnarray*}

## Ranks of pooled sample

- Let $Y_{ij}$, $i=1,\ldots, n_j$ be observations from two treatment groups $j=1,2$.
- They can also be represened by $Z_1,\ldots, Z_n$ ($n=n_1+n_2$), the outcomes of the pooled sample

```{r}
t(chol)
z=chol$cholest
z
rank(z)
```

# Wilcoxon-Mann-Whitney Test

 Simultaneously developed by Wilcoxon, and,  Mann and Whitney:  **Wilcoxon-Mann-Whitney**, **Wilcoxon rank sum test**  or **Mann-Whitney U test**

## hypotheses
  
Under $H_0$ the distributions of the two groups are equal 
$$H_0: f_1=f_2$$
  

Under the alternative $H_1$ the distributions differ in location $$H_1: \mu_1\neq \mu_2$$
  
$H_1$ assumes **location-shift**, we will relax this assumption later on.  

## Test statistic

Classic T-test: difference in sample means $\bar{Y}_1-\bar{Y}_2$.

Here: Difference in sample means based on rank transformed data 

Ranks based on the poolde sample (upon joining the observations from the two groups): $R_{ij}=R(Y_{ij})$ is de rank of observation $Y_{ij}$ in the pooled sample. 

\[
  T = \frac{1}{n_1}\sum_{i=1}^{n_1} R(Y_{i1}) - \frac{1}{n_2}\sum_{i=1}^{n_2} R(Y_{i2}) .
\]

- Under $H_0$ we expect the average rank of the first group to be close to that of the second group so $T$ is close to zero.

- Under $H_1$ we expect the mean ranks to differ so that $T$ deviates from zero. 

- It is sufficient to only calculate 
  $$S_1=\sum_{i=1}^{n_1} R(Y_{i1})$$.

- $S_1$ is the sum of the ranks of the first group: *rank sum test*.

- This holds because 
\[
  S_1+S_2 = \text{sum of all ranks} = 1+2+\cdots + n=\frac{1}{2}n(n+1).
\]


- $S_1$ (or $S_2$) is a good test statistic

- Use permutations to determine the exact permutation distribution. (Permute the ranks between the groups) 

- For a given $n$ and no *ties* the rank transformed data is always 
  $$1, 2, \ldots, n$$ 
- For given $n_1$ en $n_2$ the permutation distribution is always the same! 
- With current computing power this is not so important any more. 

## Standardized statistic 

Often the standardized test statistic is used
\[
  T = \frac{S_1-\text{E}_{0}\left[S_1\right]}{\sqrt{\text{Var}_{0}\left[S_1\right]}},
\]

- with $\text{E}_{0}\left[S_1\right]$ and $\text{Var}_{0}\left[S_1\right]$ the expect mean and variance of S1 under $H_0$. 

- Under $H_0$
 \[
   \text{E}_{0}\left[S_1\right]= \frac{1}{2}n_1(n+1) \;\;\;\;\text{ en }\;\;\;\; \text{Var}_{0}\left[S_1\right]=\frac{1}{12}n_1n_2(n+1).
 \]

- Under $H_0$ and when $\min(n_1,n_2)\rightarrow \infty$
 \[
    T = \frac{S_1-\text{E}_{0}\left[S_1\right]}{\sqrt{\text{Var}_{0}\left[S_1\right]}} \rightarrow N(0,1).  
 \]

Asymptotically the standardised statistic follows a standard normal distribution! 

## Cholestorol example 

We illustrate the result for the cholestorol example using the R function `wilcox.test`. 
```{r}
wilcox.test(cholest~group,data=chol)
```

- We reject $H_0$ ($p=$ `r format(wilcox.test(cholest~group,data=chol)$p.value,digits=2)` $<0.05$)

- The output shows $W=$ `r wilcox.test(cholest~group,data=chol)$statistic`? 

- Lets calculate 
```{r}
S1 <- sum(rank(chol$cholest)[chol$group==1])
S1
S2 <- sum(rank(chol$cholest)[chol$group==2]) 
S2
```

- Where does $W=$ `r wilcox.test(cholest~group,data=chol)$statistic` comes from? 

## Mann and Whitney test

Mann and Whitney test in absence of ties:
\[
 U_1 = \sum_{i=1}^{n_1}\sum_{k=1}^{n_2} \text{I}\left\{Y_{i1}\geq Y_{k2}\right\}.
\]

- with $\text{I}\left\{.\right\}$ an indicator that equals 1  if the expression is true and is zero otherwise. 

- U counts how many times an observation of the first group is larger or equal to an observation from the second group.

```{r}
y1=subset(chol,group==1)$cholest
y2=subset(chol,group==2)$cholest
u1Hlp=sapply(y1,function(y1i,y2) {y1i>=y2},y2=y2)
colnames(u1Hlp)=y1;rownames(u1Hlp)=y2
```

```{r}
u1Hlp
U1=sum(u1Hlp); U1
```

It can be shown that $U_1 = S_1 - \frac{1}{2}n_1(n_1+1).$

```{r}
S1-nGroups[1]*(nGroups[1]+1)/2
```

1. $U_1$ en $S_1$ contain the same information 
2. $U_1$ is also a rankstatistiek, and 
3. Exact testen based on $U_1$ and $S_1$ are equivalent. 

## Probabilistic index

- $U_1$ has a better interpretation feature
- Let $Y_j$ a random observation from group $j$ ($j=1,2$). Then 
\begin{eqnarray*}
  \frac{1}{n_1n_2}\text{E}\left[U_1\right] 
     &=& \text{P}\left[Y_1 \geq Y_2\right].
\end{eqnarray*}

So we can estimate the probability by calculating the mean of all indicator variable values $\text{I}\left\{Y_{i1}\geq Y_{k2}\right\}$. Note, that we did $n_1 \times n_2$ comparisons

```{r}
mean(u1Hlp)
U1/(nGroups[1]*nGroups[2])
```

- Probability $\text{P}\left[Y_1 \geq Y_2\right]$ is referred to as the *probabilistic index*. 
- It is the probability that a random observation of the first group is larger or equal than a random observation of the second group  
- If $H_0$ holds $\text{P}\left[Y_1 \geq Y_2\right]=\frac{1}{2}$.

- R function `wilcox.test` does not return the Wilcoxon rank sum statistic. It returns the Mann-Whitney statistiek $U_1$.
- Lets revisit the result 
```{r}
wTest<-wilcox.test(cholest~group,data=chol) 
wTest
U1
probInd=wTest$statistic/prod(nGroups) 
probInd
```

Because $p=$ `r format(wTest$p.value,digits=3)` $<0.05$ we conclude at the $5\%$ significance level that the mean cholestorol level of hartpatients is larger then that of healthy subjects. 

  - Note that we have assumed that the location-shift model is valid in this conclusion. 
  - We also know that higher cholestorol level are more likely for hartpatients then for healthy subjects and this probability is 
$U1/(n_1\times n_2)=$ `r probInd*100`%. 
  - We should assess the location shift assumption. But this is not possible with only 5 observations. 

Without the location-shift assumption the conclusion in terms of the probabilistic index remains valid! 

  - So when we do not assume location shift we test for 

\[H_0: F_1=F_2 \text{ vs } H_1: P[Y_1 \geq Y_2] \neq 0.5.\]



## Conclusion

There is a significant difference in the distribution of the cholestorol concentration of hartpatients two days upon a stroke and that of healthy subject (($p=$ `r format(wTest$p.value,digits=3)`). It is more likely to observe higher cholestorol levels for hartpatients then for healthy subjects. The point estimator for this probability is `r probInd*100`%.


