---
title: 'Tutorial 2.1: hypothesis testing on the armpit dataset'
output:
  html_document:
    code_download: yes
    highlight: tango
    number_sections: yes
    theme: cosmo
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

In this tutorial, we perform a hypothesis test on the
"smelly armpit" dataset. 

## Smelly armpit dataset

Smelly armpits are not caused by sweat, itself. The smell is caused by specific micro-organisms belonging to the group of *Corynebacterium spp.* that metabolise sweat.
Another group of abundant bacteria are the *Staphylococcus spp.*, these bacteria do not metabolise sweat in smelly compounds.

The CMET-groep at Ghent University does research to on transplanting the armpit microbiome to save people with smelly armpits.

- Proposed Therapy:
  	1. Remove armpit-microbiome with antibiotics
    2. Influence armpit microbiome with microbial transplant (https://youtu.be/9RIFyqLXdVw)

- Experiment:

    - 20 students with smelly armpits are attributed to one of two treatment groups
    - placebo (only antibiotics)
    - transplant (antibiotica followed by microbial transplant).
    - The microbiome is sampled 6 weeks upon the treatment
    - The relative abundance of *Staphylococcus spp.* on *Corynebacterium spp.* +
      *Staphylococcus spp.* in the microbiome is measured via DGGE (*Denaturing Gradient
      Gel Electrophoresis*).
    
## Goal

We find whether or not the relative abundance *Staphylococcus spp.* 
in the microbiome of the armpit is significantly different between
subjects treated with antibiotics only and subjects treated with
antibiotics and a microbial transplant.

More formally, we want to test the hypothesis that the average 
relative abundance of *Staphylococcus spp.* is equal (or not)
between both treatment groups.

Load the libraries

```{r, message=FALSE}
library(tidyverse)
```

## Import the dataset

```{r}
ap <- read_csv("https://raw.githubusercontent.com/GTPB/PSLS20/master/data/armpit.csv")
```

```{r}
glimpse(ap)
```

## Data visualization

It is always a good idea to first have a quick look at the raw data;

```{r}
ap %>% ggplot(aes(x=trt,y=rel,fill=trt)) + 
  geom_boxplot(outlier.shape=NA) + 
  geom_point(position="jitter") +
  ylab("relative abundance (%)") +
  xlab("treatment group") + 
  stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black")
```

We clearly see that, on average, the subjects who had a
microbial transplant have a higher relative abundance of
Staphylococcus spp. But is this difference **significant**?

We can test this with an unpaired, two-sample t-test. 
Btu before we can start the analysis, we must check if
all assumptions to perform a t-test are met.

## Check the assumptions

1. The observations are independent of each other (in both groups)
2. The data (rel) must be normally distributed (in both groups)
3. The variability within both groups is similar

The first assumption is met, as we may assume that there are no
specific patterns of correlation in our group of 20 randomly
select subjects. Note, however, that as we sample only in the
population of 'students', we will only be able to extrapolate
our findings to this group.

To check the normality assumption, we will use QQ plots.

```{r}
placebo_rel <- ap %>%
  filter(trt == "placebo") %>%
  pull(rel)

transplant_rel <- ap %>%
  filter(trt == "transplant") %>%
  pull(rel)

car::qqPlot(placebo_rel, main="QQ-plot of placebo")
car::qqPlot(transplant_rel, main="QQ-plot of transplants")
```

We can see that all of the data falls nicely within
the 95% confidence interval around the quantile-quantile
line (blue dashed line). As such, we may conclude that 
our data is normally distributed.

For the third assumption, we must compare the within-group
variability of both groups. We can do this visually:

```{r}
ap %>%  ggplot(aes(x=trt,y=rel)) + 
  geom_boxplot(outlier.shape=NA) + 
  geom_point(position="jitter") +
  ylab("relative abundance (%)") +
  xlab("treatment group") + 
  stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black")
```

As a measure of variability, we may take the height
of each boxplot's box. This is the interval between
the 25% and 75% quantile. Here we can see that this
interval, as well as the length of the whiskers, is
approximately equal for groups. We speak about deviation
from equality if one height is more than 2 or 3 times
larger/smaller than that of the other group.

As all three assumptions are met we may continue with
performing the unpaired two-sample t-test.

## Two-sample t-test (unpaired)

```{r}
output <- t.test(placebo_rel,transplant_rel,conf.level = 0.95,var.equal = TRUE)
output
```

Let's interpret the output.

The alternative hypothesis HA of the test is that
the "true difference in means is not equal to 0". The
null hypothesis is that the true difference in means is
equal to 0. We have found a p-value of 8.638e-05, which
is smaller than 0.05. As such, we will reject the null
hypothesis at the 5% significanc level in favor of HA.

The p-value was calculated from the t-test statistic value 
-5.0334 on a t-distribution with 18 (10+10-2) degrees of
freedom.

The mean relative abundance in the placebo group is 44%, 
while the mean relative abundance in the transplant group
is 66%. With 95% probability, the interval [-31.4; -13.0]
percentage holds the true difference in population means of 
relative abundances between subjects treated with placebo 
and subjects treated with a microbial transplant.

** Conclusion **

We may conclude that, on the 5% significance level, the mean 
relative abundance in subjects that had a placebo treatment is 
significantly (p = `r round(output$p.value,5)`) lower than the mean relative 
abundance in subjects that had a microbial transplant. The 
relative abundances are on average 
`r round( unname(output$estimate[2]) - unname(output$estimate[1]),2)` percent
(95% CI: [ `r round(output$conf.int[c(1,2)],2)` ]) lower with placebo 
treatment than with the transplant.





