---
title: 'Tutorial 2.1: Hypothesis testing on the armpit dataset'
output:
  html_document:
    code_download: yes
    highlight: tango
    number_sections: yes
    theme: cosmo
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

In this tutorial, we perform a hypothesis test on the
"smelly armpit" dataset. 

## Smelly armpit dataset

Smelly armpits are not caused by sweat, itself. The smell is caused by specific micro-organisms belonging to the group of *Corynebacterium spp.* that metabolise sweat.
Another group of abundant bacteria are the *Staphylococcus spp.*, these bacteria do not metabolise sweat in smelly compounds.

The CMET-groep at Ghent University does research to on transplanting the armpit microbiome to save people with smelly armpits.

- Proposed Therapy:
  	1. Remove armpit-microbiome with antibiotics
    2. Influence armpit microbiome with microbial transplant (https://youtu.be/9RIFyqLXdVw)

- Experiment:

    - 20 students with smelly armpits are attributed to one of two treatment groups
    - placebo (only antibiotics)
    - transplant (antibiotica followed by microbial transplant).
    - The microbiome is sampled 6 weeks upon the treatment
    - The relative abundance of *Staphylococcus spp.* on *Corynebacterium spp.* +
      *Staphylococcus spp.* in the microbiome is measured via DGGE (*Denaturing Gradient
      Gel Electrophoresis*).
    
## Goal

We find whether or not the relative abundance *Staphylococcus spp.* 
in the microbiome of the armpit is significantly different between
subjects treated with antibiotics only and subjects treated with
antibiotics and a microbial transplant.

More formally, we want to test the hypothesis that the average 
relative abundance of *Staphylococcus spp.* is equal (or not)
between both treatment groups.

Load the libraries

```{r, message=FALSE}

```

## Import the dataset

```{r}

```

Have a look at how he dataset is built up.

```{r}

```

## Data visualization

It is always a good idea to first have a quick look at the raw data;
After creating an informative visualization, do not forget your plot!

```{r}

```



## Check the assumptions

Prior to performing the t-test, we should check if all the required assumptions are met.

Check all the required assumptions. Discuss. May we continue with analysing the
data with a t-test?


## Analysis

Based on the diagnostics on the assumptions and the experimental design,
perform the appropriate test. Provide an initial interpretation of the output.

```{r}

```


## Conclusion

Writen a conclusion that is accurate, detailed and concise.
In stead of typing the required output values yourself, you 
could try to refer to an R object. For instance, given that you
stored the output of the test in an object `output`,
you could retrieve the p-value of the test as `r round(output$p.value,5)`
(rounded on 5 numbers).






