---
title: "Tutorial 2.1: Hypothesis testing on the captopril dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

In this second tutorial, we will see some basics
in hypothesis testing and more specifically on
t-tests. As an example, we will work with the
captopril dataset that we explored in the tutorial on data exploration.

We will work around these two research questions;

1) Is the average systolic blood pressure
before captopril treatment (SBPb) higher than 149 mmHg?

2) Is the average SBP before captopril treatment 
significantly different from the average SBP after
captopril treatment?

First, we will load the required R libraries:

```{r, message=FALSE}

```

# Import the data

```{r}
captopril <- ...
```

# Data exploration

**Note:** you may copy the results from the data exploration tutorial.

Before we start with hypothesis testing, it is crucial to first explore the data. 

```{r}

```

We have 15 patients, for which we have measured the systolic 
blood pressure and diastolyic blood pressure, before and after
treatment with the captopril drug.

Visualize the data in an informative way (see tutorial of yesterday)

```{r}

```

Interpret the plot!

# Question 1

Is the average systolic blood pressure (SBP)
before captopril treatment higher than 149 mmHg?

In the data exploration of the NHANES dataset we have set up a reference
interval, i.e. an interval that is expected to hold 95% of
the SBP values of healthy individuals. We found the interval
of [93;149] mmHg. (If you did not get to this point of the script
yet, don't worry!).

To test the effect of the captopril on subjects with hypertension
(patients), we need to find a group of patients that have
elevated SBP levels, higher than 149 mmHg.Therefore, we want
to test whether the patients in the captopril study indeed have on average a SBP level
that is greater than 149 mmHg. 
We can assess this research hypothesis using a one sample t-test.

## Assess the assumptions

Before we can perform a t-test, we must check that the required
assumptions are met!

1. The observations are independent of each other
2. The data (SBPb) must be normally distributed


```{r}
... %>%
  ... +
  geom_qq() +
  geom_qq_line()
```

Interpret the qq-plot

If you feel comfortable with assuming normality based on the qq-plot,
you may proceed with the analysis. 

## Hypothesis test

Here, we will 
test if mean SBPb is significantly higher than 149 mmHg.

More specifically, we will test the null hypothesis;

H0: the mean SBPb is equal to 149 mmHg

versus the alternative hypothesis;

HA: the mean SBPb is greater than 149 mmHg

```{r}
output1 <- t.test(...,mu=...,alternative = ...,conf.level = ...,data=...)
output1
```

When writing a conclusion on your research hypothesis,
it is very important to be precise and concise, yet complete.
Formulate a proper conclusion.

# Question 2

Is the average SBP before captopril treatment is different from the 
average SBP after captopril treatment?

The data are paired data; each patient's BP values were measured twice. 
Therefore, there will be a strong correlation between the BP values before and
after treatment of each individual patient. We can show this
with a scatterplot.

```{r}
# Create scatterplot
```

We clearly see that if a patient's SBPb value is high, its
SBPa value will be comparatively high as well. 

As such, we will now need to perform a paired two-sample t-test. First, we must 
check the assumptions.

## Check the assumptions

State the assumptions that you have to check and include the diagnostic plots to assess the assumption. 

## Hypothesis test

If all assumptions are met, we may continue with
performing the paired two-sample t-test.

```{r}
output2 <- t.test(...,..., paired = ...,data=)
output2
```

Formulate a conclusion bases on the output


## One-sample t-test on the difference

Performing a paired two-sample t-test is
analogous to performing a one-sample t-test on the difference
between both groups.

This can be spotted in the output of the paired two-sample
t-test. The alternative hypothesis HA there states that 
the "true difference in means is not equal to 0". So internally,
R will actually perform a one-sample t-test on the difference, and
check whether or not the true mean difference is equal to 0.
We can also set this up manually.

```{r}
captopril %>%
  mutate(bp_diff = ...) %>%
  t.test(...,mu=...)
```

Indeed, the output is completely analogous to that
of the paired two-sample t-test.











































