---
title: "Tutorial 2.2: Hypothesis testing on the captopril dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

In this second tutorial, we will see some basics
in hypothesis testing and more specifically on
t-tests. As an example, we will work with the
captopril dataset that we explored yesterday.

We will work around these two research questions;

1) Is the average systolic blood pressure
before captopril treatment (SBPb) higher than 149 mmHg?

2) Is the average SBP before captopril treatment 
significantly different from the average SBP after
captopril treatment?

First, we will load the required R libraries:

```{r, message=FALSE}

```

## Import the data

```{r}
captopril <- ...
```

## Data exploration

**Note:** you may copy the results from the data exploration tutorial.

Before we start delving into the data in order to solve 
our research hypothese, it is crucial to first look at the data. 

```{r}

```

We have 15 patients, for which we have measured the systolic 
blood pressure and diastolyic blood pressure, before and after
treatment with the captopril drug.

Visualize the data in an informative way (see tutorial of yesterday)

```{r}

```

Interpret the plot!

## Question 1

Is the average systolic blood pressure (SBP)
before captopril treatment higher than 149 mmHg?

Yesterday, we used the NHANES dataset to set up a reference
interval, i.e. an interval that is expected to hold 95% of
the SBP values of healthy individuals. We found the interval
of [93;149] mmHg. (If you did not get to this point of the script
yesterday, don't worry!).

To test the effect of the captopril on diseased individuals
(patients), we need to find a group of patients that have
elevated SBP levels, higher than 149 mmHg.Therefore, we want
to test whether or not the provided patients have an SBP level
that is greater than 149 mmHg. To test this, we will perform a
one-sample t-test, which will tell us if the average SBPb of
our patients is significantly greater than 149 mmHg on the 5%
significance level.

### Assess the assumptions

Before we can perform a t-test, we must check that the required
assumptions are met!

1. The observations are independent of each other
2. The data (SBPb) must be normally distributed


```{r}
... %>%
  ... +
  geom_qq() +
  geom_qq_line()
```

Interpret the qq-plot

If you feel comfortable with assuming normality based on the qq-plot,
you may proceed with the analysis. Here, we will 
test if mean SBPb is significantly higher than 149 mmHg.

More specifically, we will test the null hypothesis;

H0: the mean SBPb is equal to 149 mmHg

versus the alternative hypothesis;

HA: the mean SBPb is greater than 149 mmHg

```{r}
output1 <- t.test(...,mu=...,alternative = ...,conf.level = ...,data=...)
output1
```

When writing a conclusion on your research hypothesis,
it is very important to be precise and concise, yet complete.
Formulate a proper conclusion.

## Question 2

Is the average SBP before captopril treatment is different from the 
average SBP after captopril treatment?

The data are paired data; each patient's BP values were measured twice. 
Therefore, there will be a strong correlation between the BP values before and
after treatment of each individual patient. We can show this
with a scatterplot.

```{r}
## Create scatterplot
```

We clearly see that if a patient's SBPb value is high, its
SBPa value will be comparatively high as well. 

As such, we will now need to perform a paired, two-sample t-test. First, we must 
check the assumptions.

### Check the assumptions

If both assumptions are met, we may continue with
performing the paired two-sample t-test.

```{r}
output2 <- t.test(...,..., paired = ...,data=)
output2
```

Formulate a nice conclusion for the output

### One-sample t-test on the difference

Performing a paired two-sample t-test is
analogous to performing a one-sample t-test on the difference
between both groups.

This can be easily seen from the output of the paired two-sample
t-test. The alternative hypothesis HA there states that 
the "true difference in means is not equal to 0". So internally,
R will actually perform a one-sample t-test on the difference, and
check whether or not the true mean difference is equal to 0.
We can also set this up manually.

```{r}
captopril %>%
  mutate(bp_diff = ...) %>%
  t.test(...,mu=...)
```

Indeed, the output is completely analogous to that
of the paired two-sample t-test.











































