---
title: "Tutorial 2.3: Hypothesis testing on the shrimps dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

To experiment with the concepts of non-paramteric,
data analysis, we will work with the shrimps dataset.

## The shrimps dataset

A dataset on PCBs (Polychlorinated biphenyls) accumulation
in the adipose tissue of shrimps.

PCBs are often present in coolants, and are know to accumulate
easily in the adipose tissue of shrimps. In this experiment, two
groups of 18 samples (each 100 grams) of shrimps each were cultivated 
in different conditions, one control condition and one condition 
where the medium was poluted with PCBs.

Load the required R libraries

```{r}
library(tidyverse)
```

## Import the data

## Initial exploration of the data


## Goal

The research question is; is there an effect of the 
growth condition on the PCB concentration in the adipose 
tissue of shrimps. Note that the PCB concentrations were 
measured in pg/g adipose tissue.

## Data visualization

```{r}

```


We can see that for group 1 we have 4 very clear outliers
in the data. These values were double-checked (i.e for
typing errors), but there was no reason found to believe
that these values are incorrect.

As we have seen in tutorials 2.1 and 2.2, a good way for
testing the research hypothesis is to perform an unpaired
two-sample t-test to find out whether there is a significant
difference in the mean PCB concentrations between both groups
of samples. Before we can do this, we must check if all the 
required assumptions are met.

1. The observations are independent of each other (in both groups)
2. The data (PCB.conc) must be normally distributed (in both groups)

Check the assumptions

```{r}

```

Are you comfortable in taking the required assumptions?
If yes, you may proceed with the further analysis (i.e.,
perform the appropriate test and formulate a conclusion
that answers the research question).

```{r}

```


















