---
title: "7. Analysis of variance" 
author: "Lieven Clement"
date: "statOmics, Ghent University (https://statomics.github.io)"
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, comment = NA, echo = TRUE,
                      message = FALSE, warning = FALSE)
library(tidyverse)
```

# Prostacyclin Example

Researchers study the effect of arachidonic acid on prostacyclin level in blood plasma. They use 3 different concentrations of arachidonic acid: 

- low, 
- medium and 
- high dose

Each treatment is adopted to 12 rats. They measure the prostacyclin levels in blood plasma using an ELOSA fluorescence measurement.


```{r}
prostacyclin <- read_tsv("https://raw.githubusercontent.com/GTPB/PSLS20/master/data/prostacyclin.txt")
```

```{r}
head(prostacyclin)
```

```{r}
prostacyclin <- prostacyclin %>%
  mutate(dose = as.factor(dose))
```


##Data exploration

We assess if distribution of the prostacyclin concentration for each of the arachidonic acid treatments using a boxplot. We also assess if the data are normally distributed in each treatment group using QQ-plots. 

```{r}
prostacyclin %>% 
  ggplot(aes(x=dose,y=prostac,fill=dose)) +
  geom_boxplot() +
  geom_point(position="jitter") +
  ylab("prostacyclin (ng/ml)")
  
prostacyclin %>% 
  ggplot(aes(sample=prostac)) +
  geom_qq() +
  geom_qq_line() + 
  facet_grid(~dose)
```


The data in the three groups are approximately Normally distributed with equal variance:
\[Y_i \vert \text{group j} \sim N(\mu_j,\sigma^2),\]
with $j= \text{1, 2, 3}$

## Research question

Research question can translated in the following hypotheses 

- $H_0$: the arachidonic acid concentration has no effect on the mean prostacyclin level in blood plasma in rats
\[
  H_0:\mu_1=\mu_2 = \mu_3
\]

- $H_1$: the arachidonic acid concentration has an effect on the mean prostacyclin level in blood plasma in rats, which implies that the at least two means are different.

In terms of the model parameters this becomes

\[
  H_0:\mu_1=\mu_2 = \mu_3
\]
and
\[H_1: \exists\ j,k \in \{1,\ldots,g\} : \mu_j\neq\mu_k\]

We can assess this omnibus hypothesis using ANalysis of VAriance (ANOVA).

# Analyse of Variance

```{r}
model1 <- lm(prostac~dose,data=prostacyclin)
anova(model1)
```

Based on the anova analysis we can reject the omnibus null hypothesis that there is no effect of the arachidonic acid treatment on the average prostacyclin concentration in rats. 
Next we continue to assess for which treatment groups the means are different using a posthoc analysis with the Tukey method to correct for multiple testing. 

```{r}
#install.packages("multcomp")
library(multcomp)
model1.mcp<-glht(model1,linfct=mcp(dose="Tukey"))
summary(model1.mcp)
confint(model1.mcp)
plot(model1.mcp)
```

# Conclusion

- There is an extreme significant effect of arachidonic acid on the average prostacyclin blood concentration in rats ($p<0.001$).
The average prostacyclin concentration is higher in the high dose group than in the low and moderate dose group (both p-values are smaller than $p<0.001$).
- The average concentration in the high dose group is `r round(confint(model1.mcp)$confint[2,1],1)`ng/ml (95% CI [`r paste(round(confint(model1.mcp)$confint[2,2:3],1),collapse=",")`]ng/ml) and `r round(confint(model1.mcp)$confint[3,1],1)`ng/ml (95% BI [`r paste(round(confint(model1.mcp)$confint[3,2:3],1),collapse=",")`]ng/ml) higher than in the low and middle dose group, respectively.
- The difference in average prostacyclin concentration between the moderate and low dose group is not significant  (p=`r round(summary(model1.mcp)$test$pvalues[1],2)`).
(All p-values and confidence intervals for post-hoc tests are corrected for multiple testing using the Tukey method).