---
title: "Tutorial 1.2: Exploring the captopril dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

In this tutorial, you will learn how to import, tidy, wrangle and 
visualize data yourself! You will work with one specific dataset;

# The captopril dataset

The captopril dataset holds information on a small experiment with
15 patients that have increased blood pressure values. 
More specifically, for each patient we will have four values;
one value for systolic blood pressure and one for diastolyic,
both before and after treating the patient with a drug named captopril.


# Import the data


```{r, message=FALSE}
library(tidyverse)
```

```{r}
captopril <- read_csv("https://raw.githubusercontent.com/GTPB/PSLS20/master/data/captopril.txt")
```

```{r}
head(captopril)
```

** Why is this dataset not tidy? **

# Data visualization

Let's say we now first want to visualize the data. 
One possibility to easily visualize the four types
of blood pressure values is by adopting the `gather`
function from tidyverse. It will reshape the dataframe,
such that we have have a single variable `type`, which 
points at one of the four blood pressure types, and `bp`,
which points at the actual value for each type 
for each patient.

```{r}
captopril %>% 
  gather(type,bp,-id)
```

Again, not all visualization types will be equally informative.
Let us first make a barplot of the data.

```{r}
captopril %>% 
  gather(type,bp,-id) %>% 
  group_by(type) %>%
    summarize_at("bp",list(mean=mean,
              sd=sd,
              n=length)) %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(x=type,y=mean,fill=type)) + 
  scale_fill_brewer(palette="RdGy") +
  theme_bw() +
  geom_bar(stat="identity") + 
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se),width=.2) +
  ggtitle("Barplot of different blood pressure measures") +
  ylab("blood pressure (mmHg)")
```

This barplot is not very informative. De height of the bars
only provides us with information of the mean blood pressure.
However, we don't see the actual underlying values, so we for
instance don't have any information on the spread of the data.
It is usually more informative to represent to underlying 
values as _raw_ as possible. Boxplots are ideal for this!

Note that it in theory it is possible to add the raw data on
the barplot, but we still would not see any measures of the
spread, such as the interquartile range.

```{r}
captopril %>% 
  gather(type,bp,-id) %>% 
  ggplot(aes(x=type,y=bp,fill=type)) + 
  scale_fill_brewer(palette="RdGy") +
  theme_bw() +
  geom_boxplot(outlier.shape=NA) + 
  geom_jitter(width = 0.2) +
  ggtitle("Boxplot of different blood pressure measures") +
  ylab("blood pressure (mmHg)") + stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black")
```

With the boxplot, we get a lot of useful information.
We immediately see multiple features of the spread of
the data, such as the median, the 25% and 75% quantiles
and outliers. Since we only have 15 raw values (patients),
we can easily add them to the plot without getting messy.

In terms of interpretation, we can see that the median 
systolic and diastolic blood pressure values are lower 
after treatment with captopril than before. If we want to
have visual inference on the mean values (cfr. class on t-test),
we can add them to the plot `stat_summary` function.

An important feature of this dataset is that it contains
paired data; for each patient, we have blood pressure values
(systolic and diastolic) before and after treatment with
captopril.

We can visualize this as follows;

```{r}
captopril %>% 
  gather(type,bp,-id) %>% 
  filter(type%in%c("SBPa","SBPb")) %>%
  ggplot(aes(x=id,y=bp,color=type)) +
  geom_point() +
  scale_color_manual(values = c("darkgreen","red"))
```

We see that for all patients, the systolic blood
pressure is lower after captopril treatment than before.

Note that we could not see this from the boxplot, directly;


```{r}
captopril %>% 
  gather(type,bp,-id) %>% 
  filter(type%in%c("SBPa","SBPb")) %>%
  ggplot(aes(x=type,y=bp,fill=type)) + 
  scale_fill_brewer(palette="RdGy") +
  theme_bw() +
  geom_boxplot(outlier.shape=NA) + 
  geom_jitter(width = 0.2) +
  ggtitle("Boxplot of blood pressure measures before and after treatment") +
  ylab("blood pressure (mmHg)") + stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black")
```

Another good way to visualise the paired data is by using a line plot that is grouped by subject.
Note, that we make a factor of the type and that we choose order of the levels so that the level "b" before becomes before the level "a". Otherwise the after measurements will appear first in the plot. 

```{r}
captopril %>% 
  gather(type,bp,-id) %>%
  mutate(type=factor(type,levels=c("SBPb","SBPa","DBPb","DBPa"))) %>%
  filter(type%in%c("SBPa","SBPb")) %>%
  
  ggplot(aes(x=type,y=bp)) +
  geom_line(aes(group = id)) +
  geom_point()
```

A typical next step is to perform a test to find out
whether the mean systolic blood pressure value after
captopril treatment is _significantly_ different from
the mean systolic blood pressure value before treatment.

Analagously, we may subtract the `after` measurement from the
`before` measurement, and test if the difference between the 
two sets of values is significantly different from zero.

```{r}
captopril %>% 
  mutate(bp_diff = SBPa-SBPb) %>%
  ggplot(aes(x="",y=bp_diff)) +
  geom_boxplot(outlier.shape=NA) +
  geom_jitter(width = 0.2) + 
  ggtitle("Boxplot of the difference in blood pressure") +
  ylab("blood pressure difference (mmHg)") + stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black") +
  theme_bw()+
  ylim(-40,10) +
  geom_hline(yintercept = 0, color="red") # adds horizontal line to plot
```

The mean difference is lower than zero. It seems 
that on average the captopril treatment has lowered
the blood pressure by about 20 mmHg. 

In the next lecture we will assess if this difference is statistically significant. 
Therefore we rely on the assumption that the differences are normally distributed. 
We can check this with a QQ-plot: 

```{r}
captopril %>% 
  mutate(bp_diff = SBPa-SBPb) %>%
  select(bp_diff) %>%
  ggplot(aes(sample=bp_diff)) +
  geom_qq() +
  geom_qq_line()
```

The plot shows no large deviations of the straight line indicating that the data are approximately normally distributed. 







