---
title: "Tutorial 1.2: Exploring the armpit dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

In this tutorial, we will explore the "smelly armpit" dataset. 

## Smelly armpit dataset

Smelly armpits are not caused by sweat, itself. The smell is caused
by specific micro-organisms belonging to the group of
*Corynebacterium spp.* that metabolise sweat.
Another group of abundant bacteria are the *Staphylococcus spp.*,
these bacteria do not metabolise sweat in smelly compounds.

The CMET-groep at Ghent University does research to on transplanting the armpit microbiome to save people with smelly armpits.

- Proposed Therapy:
  	1. Remove armpit-microbiome with antibiotics
    2. Influence armpit microbiome with microbial transplant
       (https://youtu.be/9RIFyqLXdVw)

- Experiment:

    - 20 students with smelly armpits are attributed to one of 
      two treatment groups
    - placebo (only antibiotics)
    - transplant (antibiotica followed by microbial transplant).
    - The microbiome is sampled 6 weeks upon the treatment
    - The relative abundance of *Staphylococcus spp.* on
      *Corynebacterium spp.* + *Staphylococcus spp.* in the
      microbiome is measured via DGGE (*Denaturing Gradient Gel
      Electrophoresis*).

Load the libraries

```{r, message=FALSE}
library(tidyverse)
```

## Import the dataset

```{r}
ap <- read_csv("https://raw.githubusercontent.com/GTPB/PSLS20/master/data/armpit.csv")
```

```{r}
glimpse(ap)
```

## Goal

We find whether or not the relative abundance 
*Staphylococcus spp.*
in the microbiome of the armpit is significantly different between
subjects treated with antibiotics only and subjects treated with
antibiotics and a microbial transplant.

Tomorrow, we will perform a formal test, which will tell us
whether or not the average  relative abundance of 
*Staphylococcus spp.* is equal between for treatment groups.

Here, we will limit ourselves to visualizing the data.

## Data visualization

It is always a good idea to first have a quick look at the raw
data.

First, try to make a histogram of the data. Fill in the
missing parts in the chunk of code below to get a good-looking
visualization:

```{r}
give.n <- function(x){
  return(c(y = 85, label = length(x))) 
}

ap %>%  ggplot(aes(x=trt,y=rel,fill=trt)) +
  geom_boxplot(outlier.shape=NA) + 
  geom_point(position="jitter") +
  ylab("relative abundance (%)") +
  xlab("treatment group") + 
  stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black") +
  stat_summary(fun.data = give.n, geom = "text") 
```

Based on this plot, it seems that the relative abundace
is higher for subjects who had the transplants.
However, given the small sample size the histograms
are not optimally informative. A better option would be to
show the data in a boxplot:

```{r}
ap %>%  ggplot(aes(x=trt,y=rel,fill=trt)) +
  geom_boxplot(outlier.shape=NA) + 
  geom_point(position="jitter") +
  ylab("relative abundance (%)") +
  xlab("treatment group") + 
  stat_summary(fun.y=mean, geom="point", shape=5, size=3, color="black", fill="black")
```

We clearly see that, on average, the subjects who had a
microbial transplant have a higher relative abundance of
Staphylococcus spp. Tomorrow, we will learn how to formally test
whether or not this observed difference is statistically
**significant**.

## Descriptive statistics

Here, we will generate some informative descriptive statistics
for the dataset.

We first summarize the data and calculate the mean, standard
deviation, number of observations and standard error and store the
result in an object apRelSum via 'apRelSum<-`

1. We pipe the `ap` dataframe to the group_by function to group
the data by treatment trt `group_by(trt)`
2. We pipe the result to the `summarize_at` function to summarize
the "rel" variable and calculate the mean, standard deviation and
the number of observations 
3. We pipe the result to the `mutate` function to make a new
variable in the data frame `se` for which we calculate the
standard error 

```{r}
apRelSum<-ap%>%
  group_by(trt)%>%
  summarize_at("rel",
               list(mean=~mean(.,na.rm=TRUE),
                    sd=~sd(.,na.rm=TRUE),
                    n=function(x) x%>%is.na%>%`!`%>%sum)) %>%
  mutate(se=sd/sqrt(n))

apRelSum
```

This concludes the data exploration. Tomorrow, we will 
learn how to formally test whether or not this observed 
difference is statistically **significant**.





