---
title: "Tutorial 1.2: Exploring the captopril dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

In this tutorial, you will learn how to import, tidy, wrangle and 
visualize data yourself! You will work with one specific dataset;

# The captopril dataset

The captopril dataset stems from a small experiment with
15 patients with hypertension. 
For each patient systolic and diasystolic blood pressure measurements where taken before and after administering captopril.

Load the required libraries

```{r, message=FALSE}
library(tidyverse)
```

# Import the data

```{r}
... <- read_csv(...)
```

Have a first look at the data

```{r}
head(...) ## or glimpse()
```

# Data visualization

Let's say we now first want to visualize the data. 
One possibility to easily visualize the four types
of blood pressure values is by adopting the `gather`
function from tidyverse. It will reshape the dataframe,
such that we have have a single variable `type`, which 
points at one of the four blood pressure types, and `bp`,
which points at the actual value for each type 
for each patient.

```{r}
captopril %>% 
  gather(type,bp,-id)
```

## Barplot

A barplot is a plot that you will commonly find in papers.
The code for generating such a barplot is provided below:

```{r}
captopril %>% 
  gather(type,bp,-id) %>% 
  group_by(type) %>%
     summarize_at("bp",
               list(mean=~mean(.,na.rm=TRUE),
                    sd=~sd(.,na.rm=TRUE),
                    n=function(x) x%>%is.na%>%`!`%>%sum)) %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(x=type,y=mean,fill=type)) + 
  scale_fill_brewer(palette="RdGy") +
  theme_bw() +
  geom_bar(stat="identity") + 
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se),width=.2) +
  ggtitle("Barplot of different blood pressure measures") +
  ylab("blood pressure (mmHg)")
```

A barplot, however, is not very informative. The height of the
bars only provides us with information of the mean blood pressure.
However, we don't see the actual underlying values, so we for
instance don't have any information on the spread of the data.
It is usually more informative to represent to underlying 
values as _raw_ as possible. Boxplots are ideal for this!

Note that it is possible to add the raw data on
the barplot, but we still would not see any measures of the
spread, such as the interquartile range.

Based on this critisism, can you think of a better
visualization strategy for the captopril data?

Go ahead an try out alternative strategies for generating 
the most informative plot you can think of. In addition, 
you may use the plethora of ggplot functionalities
to make it look good as well!

```{r}
## Generate an informative plot of the captopril data
...
```

After generating the plot, make sure to interpret
the visualization thoroughly. This may greatly improve
our understanding of the data, which will aid the following
data analysis.

## Paired data
An important feature of this dataset is that it contains
paired data; for each patient, we have blood pressure values
(systolic and diastolic) before and after treatment with
captopril.

It is likely that your previous visualizition does not show
that the data is paired. Can you think of a plot that shows this
information? 

**Hint:** Try to use `geom_line(aes(group = id))` for 
visualizing the paired data in the cunk of code below

```{r}

```


As an alternative to looking at the paired data, we may think 
of the data in terms of the difference between the `after` 
measurement and the `before` measurement.

Can you visualize the difference between the systolic
blood pressure levels before and after captopril treatment?

```{r}
captopril %>% 
  mutate(bp_diff = ...) %>%
  ggplot(aes(x="",y=bp_diff)) +
  ...
```

In order to assess the effect of captopril, to which
value should you compare the difference? Why?

## Assumptions: QQ-plot

In the next lecture we will assess if this difference is statistically significant. 
We will introduce a statistical test that relies on the assumption that the differences are normally distributed. 
We can check this with a QQ-plot: 

```{r}
captopril %>% 
  mutate(bp_diff = ...) %>%
  select(bp_diff) %>%
  ggplot(aes(sample=bp_diff)) +
  # generate qq plot
```

The plot shows no large deviations of the straight line indicating that the data are approximately normally distributed. 

#Descriptive statistics

Descriptive statistics for all blood pressure measurements are found below. 

```{r}
captopril %>% 
  gather(type,bp,-id) %>% 
  group_by(type) %>%
     summarize_at("bp",
               list(mean=~mean(.,na.rm=TRUE),
                    sd=~sd(.,na.rm=TRUE),
                    n=function(x) x%>%is.na%>%`!`%>%sum))%>%
  mutate(se=sd/sqrt(n))
```

It is also very useful to calculate the descriptive statistics for the systolic bloodpressure difference because we can assess the effect size of administering captopril for every patient. 

```{r}
captopril %>% 
  ...
```


# Wrap-up 

Next, time if you explore paired data, you can immediately adopt section 3.3 the line plot and the boxplot of the differences so as to acknowledge the pairing and section 3.4. 

It is not useful to report the barplot. 





