---
title: "Tutorial 1.4: Exploring the FEV dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

In this tutorial, you will learn how to import, tidy, wrangle and 
visualize data yourself! You will work with one specific dataset;

## The FEV dataset

The FEV, which is an acronym for forced expiratory volume,
is a measure of how much air a person can exhale (in liters) 
during  a forced breath. In this dataset, the FEV of 606 children,
between the ages of 6 and 17, were measured. The dataset
also provides additional information on these children:
their `age`, their `height`, their `gender` and, most
importantly, whether the child is a smoker or a non-smoker.

The goal of this experiment was to find out whether or not
smoking has an effect on the FEV of children.

Note: to analyse this dataset properly, we will need some
relatively advanced modeling techniques. At the end of this 
week, you will have seen all three required steps to analyse
such a dataset! For now, we will limit ourselves to exploring
the data.

## Load the required libraries

```{r, message = FALSE}
...
```

## Import the data

```{r}
...
```

Have a first look at the data

```{r}
...
```


There are a few things in the formatting of the
data that can be improved upon:

1. Both the `gender` and `smoking` can be transformed to
factors.
2. The `height` variable is written in inches. Assuming that
this audience is mainly Portuguese/Belgian, inches are hard to
interpret. Let's add a new column, `height_cm`, with the values
converted to centimeters

```{r}
...
```

That's better!

Now, let's make a first explorative plot, showing
only the FEV for both smoking categories.

Which type of plot do you suggest? Generate a good-looking,
informative representation of the data.

```{r}
...
```

Did you expect these results?

If your visualization is informative, it should appear that
children that smoke have a higher median FEV than children 
that do not smoke. Should we change legistlations worldwide and
make smoking obligatory for children?

Maybe there is something else going on in the data. By taking
more of the information in the dataset into account, can
you provided a more detailed/accurate visualizition of the
variables that effect the FEV?

```{r}
...


## Try to get a visualization that describes the data as good as possible!!



...
```












