---
title: "Tutorial 1.3: Exploring the captopril dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

In this tutorial, you will learn how to explore and summarize data that are paired.

# The captopril dataset

The captopril dataset stems from a small experiment with
15 patients with hypertension. For each patient systolic and
diasystolic blood pressure measurements where taken before and
after administering captopril.

Before we can start visualizing the data, we must load the
required libraries.

```{r, message=FALSE}
library(tidyverse)
```

# Import the data

```{r}

```

Have a first look at the data

```{r}

```

# Data visualization

Let's say we now first want to visualize the data. 
One possibility to easily visualize the four types
of blood pressure values is by adopting the `gather`
function from tidyverse. It will reshape the dataframe,
such that we have have a single variable `type`, which 
points at one of the four blood pressure types, and `bp`,
which points at the actual value for each type 
for each patient.

```{r}
captopril %>% 
  gather(type,bp,-id)
```

## Barplot

A barplot is a plot that you will commonly find in papers.
The code for generating such a barplot is provided below:

```{r}
captopril %>% 
  gather(type,bp,-id) %>% 
  group_by(type) %>%
     summarize_at("bp",
               list(mean=~mean(.,na.rm=TRUE),
                    sd=~sd(.,na.rm=TRUE),
                    n=function(x) x%>%is.na%>%`!`%>%sum)) %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(x=type,y=mean,fill=type)) + 
  scale_fill_brewer(palette="RdGy") +
  theme_bw() +
  geom_bar(stat="identity") + 
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se),width=.2) +
  ggtitle("Barplot of different blood pressure measures") +
  ylab("blood pressure (mmHg)")
```

A barplot, however, is not very informative. The height of the
bars only provides us with information of the mean blood pressure.
However, we don't see the actual underlying values, so we for
instance don't have any information on the spread of the data.
It is usually more informative to represent to underlying 
values as _raw_ as possible. Note that it is possible to add the
raw data on the barplot, but we still would not see any measures
of the spread, such as the interquartile range.

Based on this critisism, can you think of a better
visualization strategy for the captopril data?

# Descriptive statistics

- Provide a code chunk to calculate useful summary statistics for the capropril data!

