---
title: "Tutorial 7.2: Multiple regression on the fish tank dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

As an exercise on multiple regression, we will analyse
the fish tank dataset.

## Fish tank dataset

In this experiments 96 fish (dojofish, goldfish and zebrafish)
were placed separately in a tank with two liters of water and
a certain dose (in mg) of a certain poison EI-43,064. The resistance
of the fish a against the poison was measured as the amount of
minutes the fish survived upon adding the poison (Surv_time, in
minutes). Additionally, the weightt of each fish was measured.

## Goal

In this tutorial, we will study the association between dose and
survival time, while correcting for weight and species, by using 
a multiple regression model.

Read the required libraries

```{r, message = FALSE}
library(readr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(car)
```


## Import the data

```{r, message=FALSE}
poison <- read_csv("poison.csv")
```

## Data tidying

```{r}
head(poison)
```

We can see a couple of things in the data that can
be improved upon:

1. Capitalize the fist column name 
2. Set the Species column as a factor
3. Change the speciec factor levels from 0, 1 and 2 to
Dojofish, Goldfish and Zebrafish. Hint: use the fct_recode
function.
4. Add the variable log.Surv_time: we already saw in previous
tutorials that this transfromation is required to obtain
normally distributed data.

```{r}
poison <- poison %>%
  rename("Species" = "species") %>%
  mutate(Species = as.factor(Species)) %>%
  mutate(Species = fct_recode(Species, Dojofish = "0", Goldfish = "1", Zebrafish = "2")) %>%
  mutate(log.Surv_time = log(Surv_time))

poison
```

## Data exploration

In a previous tutorial, we already studied the effect of
dose on (the logarithm of) suvival time. There, we did not
account for the fact that fish of the same species and/or
fish of similar weight will probably reacht similarly to
the poison.

When we expect the data more closely, we see that these
factors do indeed matter.

```{r}
plot(poison)
```


```{r}
library(GGally)
poison %>% select(-Surv_time) %>% ggpairs()
```

- Survival time seems to be strongly associated with dose, 
species and weight. 

- There is a strong, positive association between log survival time 
and weight. 

- For low weights, the log survival time slope seems to flatten. 

- The weigths are not equally distributed over the different doses.

- There is an association between species and weight.

Some additional visualizations:

- Plot the log survival time in function of weigth.
Add species as a color.

```{r}
poison %>%
  ggplot(aes(x=Weight,y=log.Surv_time)) +
  geom_point() +
  stat_smooth(method = "loess") +
  ylab("log survival time") +
  xlab("Weight (g)") +
  geom_point(aes(col=Species)) + 
  scale_color_manual(values = c("red","darkgoldenrod","black")) +
  theme_bw()
```

We observe a very strong relationship between weight and log 
survial time. The weigth is different for different species.

- Plot the fish weights in function of dose

```{r}
poison %>%
  ggplot(aes(x=as.factor(Dose),y=Weight,fill=Dose)) +
  geom_boxplot(outlier.shape=NA) + 
  geom_jitter(width = 0.2) +
  scale_fill_gradient(low = "darkgreen", high = "red") +
  ylab("Weight (g)") +
  xlab("Dose (mg)") +
  theme_bw()
```

We see that, probably by chance, some dosages were administered
to fish with slightly higher (e.g. dose 1.9 mg) or lower weights.

- Plot the log of survival time in function of dose.

```{r}
poison %>%
  ggplot(aes(x=Dose,y=log.Surv_time)) +
  stat_smooth(method = "loess") +
  ylab("log survival time") +
  xlab("Dose (mg)") +
  geom_point(aes(col=Species)) +
  scale_color_manual(values = c("red","darkgoldenrod","black")) +
  theme_bw() +
  geom_smooth(aes(group=Species,col=Species), method='lm')
```

There is a clear linear relationship between dose and
log survival time, but especially the intercept of
the linear model will be different for different species.

- Plot the relationship between log survival time and species.

```{r}
poison %>%
  ggplot(aes(x=Species,y=log.Surv_time,color=Species)) +
  geom_boxplot() +
  geom_jitter() +
  ylab("log urvival time") +
  scale_color_manual(values = c("red","darkgoldenrod","black")) +
  theme_bw()
```

The log survival time is different for different species.

The researchers assume, based on the data exploration, that the
dose effect on survival varies between different species. As such,
a good model should have an interaction effect between dose and 
species. We will also include an interaction between weight and 
species.
