---
title: "Tutorial 8.2: Multiple regression on the FEV dataset"   
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

As an exercise on multiple regression, we will analyse
the FEV dataset.

## The FEV dataset

The FEV, which is an acronym for forced expiratory volume,
is a measure of how much air a person can exhale (in liters) 
during  a forced breath. In this dataset, the FEV of 606 children,
between the ages of 6 and 17, were measured. The dataset
also provides additional information on these children:
their `age`, their `height`, their `gender` and, most
importantly, whether the child is a smoker or a non-smoker.

The goal of this experiment was to find out whether or not
smoking has an effect on the FEV of children.

# Load the required libraries

```{r, message = FALSE}
library(tidyverse)
library(GGally)
```

# Import the data

```{r}
fev <- read_tsv("https://raw.githubusercontent.com/GTPB/PSLS20/master/data/fev.txt")
```

```{r}
head(fev)
```

# Data tidying

There are a few things in the formatting of the
data that can be improved upon:

1. Both the `gender` and `smoking` can be transformed to
factors.
2. The `height` variable is written in inches. Assuming that
this audience is mainly Portuguese/Belgian, inches are hard to
interpret. Let's add a new column, `height_cm`, with the values
converted to centimeters

```{r}
fev <- fev %>%
  mutate(gender = as.factor(gender)) %>%
  mutate(smoking = as.factor(smoking)) %>%
  mutate(height_cm = height*2.54)

head(fev)
```

That's better!

## Data exploration

```{r}
fev$height
```


```{r}
fev %>% mutate(lfev=log(fev)) %>% dplyr::select(smoking,gender,age,height_cm,lfev) %>% ggpairs()
```

```{r}
plot(log(fev$fev)~fev$age)
```


There a very strong associations between
- age and height
- age and FEV
- heigth and FEV
- gender and height
- gender and FEV
- ...

Remember the "Data_exploration_FEV.Rmd" file? There, we
saw that plotting the FEV in function of smoking status
only, it appeared that the FEV was higher for smokers.

```{r}
fev %>%
  ggplot(aes(x=smoking,y=age,fill=smoking)) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(width = 0.2, size = 0.1, position = position_jitterdodge()) +
  theme_bw() +
  scale_fill_manual(values=c("dimgrey","firebrick")) +
  ggtitle("Boxplot of FEV versus smoking") +
  ylab("fev (l)") +
  xlab("smoking status")
```

```{r}
plot(fev$age~fev$height_cm)
```


However, if we "corrected" the visualization for a child's
age and/or heigth and/or gender, this completely changed the picture.

Let's again make a nice plot where we make a boxplot of the FEV 
in function of age (as factor), stratified on gender (facet) 
and colored based on the smoking status.

```{r}
fev %>%
  ggplot(aes(x=as.factor(age),y=fev,fill=smoking)) +
  geom_boxplot(outlier.shape=NA) +
  geom_point(width = 0.2, size = 0.1, position = position_jitterdodge()) +
  theme_bw() +
  scale_fill_manual(values=c("dimgrey","firebrick")) +
  ggtitle("Boxplot of FEV versus smoking, stratified on age and gender") +
  ylab("fev (l)") +
  xlab("age (years)") + 
  facet_grid(rows = vars(gender))
```

# Analysis

```{r}
lm_sa <- lm(log(fev)~smoking+age+gender, data=fev)

summary(lm_sa)
plot(lm_sa)
```


```{r}
table(fev$gender,fev$smoking)
```

```{r}

```


```{r}
full_formula <- log(fev)~ age + height_cm + smoking + gender + smoking:age + smoking:height_cm + smoking:gender + age:height_cm + age:gender + height_cm:gender

lm_full <- lm(full_formula, data=fev)

summary(lm_full)
```

```{r}
lm_red <- lm(log(fev)~ age + height_cm + smoking + gender, data=fev)

summary(lm_red)
```


```{r}
lm_2 <- lm(log(fev)~ age + height_cm + gender + smoking + age:height_cm + age:gender + height_cm:gender, data=fev)

anova(lm_full, lm_2)
```

```{r}
summary(lm_2)
```

```{r}
car::vif(lm_2)
```


```{r}
summary(lm_red)
```


```{r}
summary(lm_2)
```


```{r}
lm_3 <- lm(log(fev)~ age + height_cm + gender + age:height_cm + age:gender + height_cm:gender, data=fev)

anova(lm_full, lm_3)
```

```{r}
vif(lm_3)
```


```{r}
plot(lm_full)
```

```{r}
summary(lm_full)
```

```{r}
lm_2 <- lm(log(fev)~ age + height_cm + smoking + gender + smoking:age, data=fev)

plot(lm_2)

summary(lm_2)
```

```{r}
install.packages("margins")
library(margins)
```


```{r}
lm_3 <- lm(log(fev)~ age + height_cm + smoking + gender + smoking:age, data=fev_11_15)
summary(margins(lm_3))
#summary(lm_3)
```


## Multiple regression analysis with forward model selection

Now, we will perform a multiple regression analysis on the data,
accounting for all the required variables that may influence the
relationship between smoking and FEV.

** Note: to meet the required assumptions for linear regression, **
** we will need to log-tranform the FEV response variable. **

Having spurious terms in the model (main effects and interaction terms)
may make the model worse. If an effect or interaction has no added
value, it might be better not to have it in the model.

We can select the best/required terms by performing a forward selection.
Here, we will start with an _empty_ model, i.e. a model without
explanatory variables. Each step, we will add the most useful
variable to the model

```{r}
full_formula <- log(fev)~ age + height_cm + smoking + gender + smoking:age + smoking:height_cm + smoking:gender

subfev <- fev[,c(-3)] ## retain the potential explanatory variables and the  response variable, remove height (inches)

m1 <- lm(log(fev)~1, subfev) # empty model
add1(m1, scope=full_formula, data=subfev, test="F")
```

The variable _height-cm_ has the most significant contribution 
to the model (based on p-value and AIC. As such, we will include 
it in our improved model:

```{r}
m2 <- update(m1,~.+height_cm)
add1(m2, scope=full_formula, data=subfev, test="F")
```

After doing this, _age_ has the most significant contribution 
to the model. As such, we will include it in our improved model:

```{r}
m3 <- update(m2,~.+age)
add1(m3, scope=full_formula, data=subpoison, test="F")
```

After including height_cm and age in the model,
the _smoking_ status (again) appears significant.
Include it in the model:

```{r}
m4 <- update(m3,~.+smoking)
add1(m4, scope=full_formula,data=subfev, test="F")
```

After including the smoking term, none of the remaining
model parameters (including both interactions) appear
significant (5%). As such, we won't include any other
parameters in the model. As such, based on the forward
selection, we assume the following model:

$$y_i=\beta_0+\beta_a x_{ia} + \beta_h x_{ih} + \beta_{s1} x_{is1} + \epsilon_i,$$

This model has a main effect for smoking (s), age (a) and height (h).

## Check the assumptions

For checking the assumptions of the final model,
we will first fit it:

```{r}
lm_full <- lm(log(fev) ~ age + height_cm + smoking, data=fev)
plot(lm_full)
```

List assumptions:

1. The observations are independent of each other
2. Linearity between the response and predictor variable
3. The residues of the model must be normally distributed
4. Homoscedasticity of the data

1. The first assumption is **met**: children that are more 
alike in terms of age, height and smoking status will probably 
have more similar FEV values, but this is accounted for by 
the model.

2. The linearity assumption is **met** (see plot 1).

3. The normality assumption is **met** (see plot 2).

4. The homosccedasticity assumption is **met** (see plot 3).

## Conclusion

```{r}
library(car)
Anova(lm_full, type=3)
```

```{r}
coefficients(lm_full)
confint(lm_full)
```


- Age main effect: The log FEV for children is, on 
average, `r format(abs(lm_full$coefficients["age"]),digits=3)` liters
higher for every unit increase in age (year), after correcting
for height and smoking effects (95% CI [0.01725221; 0.03166769]).
This increase is extremely significant (p=6.001e-11) on the 5% 
significance level.

- Heigth main effect: The log FEV for children is, on 
average, `r format(abs(lm_full$coefficients["height_cm"]),digits=3)` 
liters higher for every unit increase in height (cm), after correcting
for age and smoking effects (95% CI [0.01561981; 0.01827859]).
This increase is extremely significant (p < 2.2e-16) on the 5% 
significance level.

- Smoking main effect: The log FEV for children is, on 
average, `r format(abs(lm_full$coefficients["smoking"]),digits=3)` 
liters lower for smokers than for non-smokers, after correcting
for age and height effects (95% CI [-0.09651575; -0.01381958]).
This decrease is highly significant (p =0.009007) on the 5% 
significance level.












